---
title: "mcData205 WeekTwo Assignment"
author: "Steve Dutky Montgomery College"
date: "2/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results = TRUE)
```

```{r eval=FALSE}

#$Log: weekTwoInClassAssignment.Rmd,v $
#Revision 1.2  2020/02/11 22:34:26  dutky
#hyperlinks inserted
#
#Revision 1.1  2020/02/11 21:53:44  dutky
#Initial revision
#

```


```{r}
library(tidyverse)
library(httr)
library(readxl)
library(pander)
```
```{r}
## Useful functions:
tSummary<-function(obj,columnNamePrefix="") {
  # returns a dataframe of summary(obj) transposed and 
  # rownames<-names(obj), names<-the name of the summary
  # statistic prepended by columnNamePrefix
  
  # coerce the transpose of the table returned by summary() to a data frame
  a<-as.data.frame.matrix(t(summary(obj)))
  
  # for each summary statistic
  for ( i in 1:ncol(a)) {
    # extract names and values from text of summary:
    statNames<-sub("^([^:]*):.*$","\\1", a[,i] )
    statNames<-sub("^([^ ]+|[^ ]+ [^ ]+) *$","\\1", statNames )
    a[,i]<- as.numeric(sub("^[^:]*: *([^ ]*) *$","\\1", a[,i] ))
    
    # choose the most frequent name of the column
    # specifically to exclude character columns
    choiceName<-names(sort(table(statNames),decreasing = TRUE))[1]
    names(a)[i]<-paste(columnNamePrefix,choiceName,sep="")
    
    # NA out any values that do not have the chosen statistic
    a[!grepl(choiceName,statNames),i]<-NA
  }
  a
}
```


```{r}
# init
# for this assignment, we're going to choose only these stats:
stats<-c("Mean","Min.","Median","Max.","NA's")
panderOptions("digits",2)
```


```{r eval=FALSE}
## [Data Readme](https://opportunityinsights.org/wp-content/uploads/2018/04/health_ineq_online_table_10_readme.pdf) :
GET(url="https://opportunityinsights.org/wp-content/uploads/2018/04/online_data_tables-2.xls",write_disk(tf <- tempfile(fileext = ".xls")))
   
# df<-read_xls(tf)
```
##  DATA 205 Dataset Feedback Form {#top}
<br>

>   Datasets Below:

*   [Household Income and Incarceration for Children from Low-Income Households by County, Race, and Gender](#ds1)
*   [Baseline Cross-Sectional Estimates by College](#ds2)
*   [Neighborhood Characteristics by County](#ds3)

### Dataset: [Household Income and Incarceration for Children from Low-Income Households by County, Race, and Gender](https://opportunityinsights.org/wp-content/uploads/2018/10/county_outcomes_simple.csv) {#ds1}
>  From [Opportunity Insights](https://opportunityinsights.org/). See [readme](https://opportunityinsights.org/wp-content/uploads/2019/07/Codebook-for-Table-2.pdf) for full descriptions of all columns. <[top](#top)>

<br>

>  The table below omits standard error and gender specific columns.  This dataset examines for 3219 localities, household income ranking (2014-2015) and incarceration fraction (as of April 10, 2010) of adults aged 31 to 37 where the household incomes of their parents were at the 25th national percentile.

```{r}
df<-read_csv("https://opportunityinsights.org/wp-content/uploads/2018/10/county_outcomes_simple.csv")
```


```{r}
cols<-c("kfr_pooled_pooled_p25", 
 "jail_pooled_pooled_p25",
 "pooled_pooled_count",
 "kfr_black_pooled_p25", 
 "kfr_hisp_pooled_p25",
 "kfr_white_pooled_p25",
 "jail_black_pooled_p25",
 "jail_hisp_pooled_p25",
 "jail_white_pooled_p25",
 "white_pooled_count",
 "black_pooled_count",
 "hisp_pooled_count")

pander(tSummary(df[,cols])[stats])
```
#### Data Quality
*    There seem to be large proportions of NA's for both Hispanics and African Americans compared to whites.
*    There is a large variation in the population size of the localities.
*    Noise was added to the data to protect individual privacy.  That may account for the negative minimum incarceration fractions.
*    The subject counts in many localities are very small making for distorted interpreations.
*    It is unclear when or how the parent's membership in the 25th household income percentile was determined.  

#### Primary Data Columns
*    All data columns appear interesting
*    The income and incarceration should be studied independently.
*    It would be interesting to drill down on the gender differences for income.

#### Relevant Filters
*    Select for localities where the pooled-pooled count exceeds the median.
*    Select gender male/female
*    Select for localities where pooled-pooled income fraction exceeds .5

#### Possible visualations
*    Scatter plot count vs incarceration rates shapes for race/gender
*    Scatter plot count vs income fraction shapes for race/gender
*    Bar graphs for income/incarceration aggregated by state stacked by race and gender

#### Limitations of the Dataset
*   Use of national income distribution may distort local results
*   As above, NAs more frequently occur with Hispanics, African Americans.
*   Unclear whether subjects reside in childhood locality.
*   Individuals often incarerated far from home.  Local population, income distorted in rural localities with high density prisons.

#### Improvement to the Dataset
*   Categorize locality by urban, suburban, rural
*   Add adjustement locality income distribution
*   Add locality incarceration rate.

#### Always interesting to see how Montgomery County stacks up nationally:

```{r}
moco<-df[df$state==24 & df$county==31,cols]
ebElse<-tSummary(df[,cols])[stats[-5]]
names(ebElse)<-paste("all.",stats[-5],sep="")
pander((cbind(moco=t(moco),ebElse)),split.tables=90)
```
#### and in the DC commuting zone

```{r}
moco<-df[df$state==24 & df$county==31,cols]
dcCz<-tSummary(df[df$cz==11304,cols])[stats[-5]]
names(dcCz)<-paste("dcCz.",stats[-5],sep="")
pander((cbind(moco=t(moco),dcCz)),split.tables=90 ,split.cells=rep(2,6))
```

### Dataset: [Baseline Cross-Sectional Estimates by College](https://opportunityinsights.org/wp-content/uploads/2018/04/mrc_table2.csv) {#ds2}

>  From [Opportunity Insights](https://opportunityinsights.org/).  See [readme](https://opportunityinsights.org/wp-content/uploads/2018/04/Codebook-MRC-Table-2.pdf) for full descriptions of all columns. <[top](#top)>
<br><br>
> The dataset consists of 2199 rows representing 2197 post-secondary academic institutions identified with the U.S. Dept of Education's Office of Postsecondary Education Identification (OPE ID) number. 
<br><br>
>  Each row has 85 columns describing the school demographics (location, name, type, tier {prestige 1-14}, etc.) and data collected/imputed for students born 1980-1982 (count,fraction female, married, measures of income, parental income, expected income rankings. 
<br><br>
>  Column prefixes: k for student (kid), par for parent, mr_k for student mobility rate.
<br>
>  Table below includes the following columns:

```{r}
a<-structure(list(`column name` = c("type", "tier", "iclevel", "count", 
"female", "k_married", "mr_kq5_pq1", "par_rank", "par_top1pc", 
"k_rank", "k_0inc", "k_q5", "k_rank_cond_parq1", "k_rank_cond_parq2", 
"k_rank_cond_parq3", "k_rank_cond_parq4", "k_rank_cond_parq5"
), description = c("Type : 1,= public 2,= private non-profit 3,= for-profit", 
"Selectivity and type combination rank 1-14", "Four-year or two-year college", 
"Average number of kids per cohort", "Fraction female among kids", 
"Fraction of kids married in 2014", "Mobility rate (joint probability of parents in bottom quintile and child in top quintile of the income distribution)", 
"Mean parental income rank", "Fraction of parents in the top 1 income percentile [PCTILE].", 
"Mean kid earnings rank", "Fraction of kids with zero labor earnings", 
"Fraction of kids in the top income quintile", "Mean kid earnings rank conditional on parent in 1st quintile", 
"Mean kid earnings rank conditional on parent in 2nd quintile", 
"Mean kid earnings rank conditional on parent in 3rd quintile", 
"Mean kid earnings rank conditional on parent in 4th quintile", 
"Mean kid earnings rank conditional on parent in 5th quintile"
)), class = c("spec_tbl_df", "tbl_df", "tbl", "data.frame"), row.names = c(NA, 
-17L), spec = structure(list(cols = list(`column name` = structure(list(), class = c("collector_character", 
"collector")), description = structure(list(), class = c("collector_character", 
"collector"))), default = structure(list(), class = c("collector_guess", 
"collector")), skip = 1), class = "col_spec"))

cols<-unlist(a[,1])
pander(a)
```



```{r}
df<-read_csv("https://opportunityinsights.org/wp-content/uploads/2018/04/mrc_table2.csv")
```


```{r}
pander(tSummary(df[,cols])[stats])
```
#### Data Quality

*   Not clear when income ranks of students and parents were recorded. Cohorts presumably graduated high school between 1998 and 2001. Certain columns indicate value as of 2014 when students were in their thirties. Others don't.
*   Data do not distinguish when or how students began or left college.
*   Cohort began careers at the onset of the Great Recession. Presumably this dispropotionately affected their incomes.

#### Primary Data Columns

> The dataset contains a large variety of columns and offers many opportunities for studying how the interact:

*   Effect of parent's income rank on the prestige (column tier) of the institution their children attend.
*   Influence of parent's income rank on their child income rank conditioned by the prestige of their institution
*   Influence of instition's type (public/private non-profit/for-profit) and iclevel (4year/2year/<2year) on student's future (?) income ranking and/or unemployment ($0 income).
*   Other effects such as those related to gender and marital status appear difficult to tease out.

#### Relevant Filters
*   Institution prestige
*   Institution type and iclevel (see above)

#### Possible visualations
*   Scatter plot of parent vs child income rank colored by institution prestige
*   Scatter plot of cohort count vs child income rank colored by institution prestige
*   Stacked bar chart of institution prestige (horizontal) vs cohort counts by type and iclevel

#### Limitations of the Dataset
*   Dollar amount columns are not corrected for inflation.  Does not permit valid comparsion of child/parent incomes
*   Dataset fails to adequately account for cohort who attended post-secondary institutions when they over 22 years old.
*   Schools of Art and Design (eg Rhode Island School Of Design) misclassified as non selective, low prestige institutions
*   U.S. military academies omitted.

#### Improvement to the Dataset
*   Include institutions' distribution STEM/arts/business/trades/military
*   Include fraction of cohort from single parent households
*   Include cohort's minority distribution
*   Include fraction of students who did not graduate

#### Always interesting to see how Montgomery College stacks up nationally:

```{r }
moco<-df[df$super_opeid==6911,cols]
ebElse<-tSummary(df[,cols])[stats[-5]]
names(ebElse)<-paste("all.",stats[-5],sep="")
pander((cbind(MC=t(moco),ebElse)),split.tables=90)
```
#### and in the DC commuting zone

```{r}
moco<-df[df$super_opeid==6911,cols]
dcCz<-tSummary(df[df$cz==11304,cols])[stats[-5]]
names(dcCz)<-paste("dcCz.",stats[-5],sep="")
pander((cbind(MC=t(moco),dcCz)),split.tables=90 ,split.cells=rep(2,6))
```

### Dataset: [Neighborhood Characteristics by County](https://opportunityinsights.org/wp-content/uploads/2018/12/cty_covariates.csv) {#ds3}

>  From [Opportunity Insights](https://opportunityinsights.org/).  See [readme](https://opportunityinsights.org/wp-content/uploads/2019/07/Codebook-for-Table-10.pdf) for full descriptions of all columns. <[top](#top)>


```{r}
df<-read_csv("https://opportunityinsights.org/wp-content/uploads/2018/12/cty_covariates.csv")
```
<br>

>   Table below shows a subset of the columns for data 2010 and after. All units are fractions of the totals with the following exceptions:

*    **med_hhinc2016:**  median household income in 2016 in dollars
*    **gsmn_math_g3_2013:** mean score on 3rd grade standardized math test
*    **rent_twobed2015:**  median rent of two bedroom dwelling in dollars
*    **mail_return_rate2010:** percentage of completed census forms returned
*    **popdensity2010:**  residents per squire mile
*    **job_density_2013:** jobs per square mile

```{r}
cols<-c("state",
 "county",
 "cz",
 "czname",
 "frac_coll_plus2010",
 "foreign_share2010",
 "med_hhinc2016",
 "poor_share2010",
 "share_white2010",
 "share_black2010",
 "share_hisp2010",
 "share_asian2010",
 "gsmn_math_g3_2013",
 "rent_twobed2015",
 "singleparent_share2010",
 "traveltime15_2010",
 "mail_return_rate2010",
 "popdensity2010",
 "ann_avg_job_growth_2004_2013",
 "job_density_2013")

pander(tSummary(df[,cols])[stats])
```
#### Data Quality

*    Across all numeric columns there is a median NA frequency of 2.4% of the number of rows.
*    Most recent data originated in 2015. Is it still relevant unless combined with more recent data?
*    Data span the Great Recession possibly affecting dollar amount columns

#### Primary Data Columns

*   Yearly data spanning different years poses difficulties for correlating them.
*   **county** and **commuting zone**: interesting to see the variations across zones comprising multiple counties, particularly those crossing state boundaries (requires additional dataset) 
*   **poor_share[1990,2000,2010]**, **singleparent_share[same range], and **popdensity[2000,2010]**

#### Relevant Filters
*   sum(abs(shareSY1-shareSY2) > threshold for all S
#### Possible visualations
*   Stacked bar chart of counties in commuting zones sized by share column of selected year and ordered by mean population density across commuting zone. Many interesting columns.
*   Paired stacked bar chart of all shares of paired years y1,y2 in same county ordered by sum of absolute share difference y2-y1.


#### Limitations of the Dataset
*   Data are not up to date
*   Data columns span different years
*   Dollar amount data columns do not use constant dollars
*   Do commuting zones change over time?

#### Improvement to the Dataset
*   Convert dollar amounts to constant dollars
*   Join county population obtained from U.S. census


#### Always interesting to see how Montgomery County stacks up nationally:

```{r}
moco<-df[df$state==24 & df$county==31,cols]
ebElse<-tSummary(df[,cols])[stats[-5]]
names(ebElse)<-paste("all.",stats[-5],sep="")
pander((cbind(moco=t(moco),ebElse)),split.tables=120)
```
#### and in the DC commuting zone

```{r}
moco<-df[df$state==24 & df$county==31,cols]
dcCz<-tSummary(df[df$cz==11304,cols])[stats[-5]]
names(dcCz)<-paste("dcCz.",stats[-5],sep="")
pander((cbind(moco=t(moco),dcCz)),split.tables=120) 
```
